---
title: "STATS 326 Assignment 2"
Student ID: 5553528
output:
  word_document: default
  html_document: default
Name: Danielle Parkinson
---

```{r}
library(readr)
library(timeSeries)
library(s20x)
```

Question 1: 


```{r}
q_dioxide.df = read.table("quarterly dioxide.txt", header = T)
QuarterlyDioxide.ts = ts(q_dioxide.df$dioxide, frequency = 4, start = 2000)
plot.ts(QuarterlyDioxide.ts, main = "Quarterly Measurements of Atmospheric Carbon Dioxide (2000 - 2017)", xlab = "Months", ylab = "Carbon Dioxide in parts per million (ppm)")

```
Comments on the plot:
comment - increasing trend and strong seasonaility
do not transform 


Question 2:

# 1 - Workings seasonal factor model
```{r}
Time = 1:67
Quarter = factor(c(rep(1:4,16),(1:3)))
carbon.ts = ts(q_dioxide.df$dioxide[1:67],start=2000, frequency=4)
carbon.fit = lm(carbon.ts~Time+Quarter)
plot.ts(residuals(carbon.fit),main="Residuals of carbon fit")

carbon.fit2 = lm(carbon.ts~Time + I(Time^2)+Quarter)
#plot.ts(residuals(carbon.fit2),main="Residuals of carbon fit2")

#summary(carbon.fit2)
#acf(residuals(carbon.fit2))
carbon.fit2lag = lm(carbon.ts[-1]~Time[-1]+I(Time[-1]^2)+Quarter[-1]+carbon.ts[-67])
plot.ts(residuals(carbon.fit2lag), main="Carbon fit quadratic with lag residuals")
acf(residuals(carbon.fit2lag))
normcheck(residuals(carbon.fit2lag),shapiro.wilk = T)
summary(carbon.fit2lag)

#Forecasting for Q4 2016
t68.pred = carbon.fit2lag$coefficients[1] + carbon.fit2lag$coefficients[2] *68 + carbon.fit2lag$coefficients[3] * (68^2) - carbon.fit2lag$coefficients[6] *1 + carbon.fit2lag$coefficients[7] * carbon.ts[67]
t68.pred

t69.pred = carbon.fit2lag$coefficients[1] + carbon.fit2lag$coefficients[2] *69 + carbon.fit2lag$coefficients[3] * (69^2) + carbon.fit2lag$coefficients[7] * t68.pred
t69.pred

t70.pred = carbon.fit2lag$coefficients[1] + carbon.fit2lag$coefficients[2] *70 + carbon.fit2lag$coefficients[3] * (70^2) - carbon.fit2lag$coefficients[4] *1 + carbon.fit2lag$coefficients[7] *t69.pred
t70.pred

t71.pred = carbon.fit2lag$coefficients[1] + carbon.fit2lag$coefficients[2] *71 + carbon.fit2lag$coefficients[3] * (71^2) - carbon.fit2lag$coefficients[5] *1 + carbon.fit2lag$coefficients[7] * t70.pred
t71.pred

print("The RMSEP")
RMSEP.seasonal_factor = sqrt((1/4)*sum((q_dioxide.df$dioxide[68]-t68.pred)^2,(q_dioxide.df$dioxide[69]-t69.pred)^2,(q_dioxide.df$dioxide[70]-t70.pred)^2,(q_dioxide.df$dioxide[71]-t71.pred)^2))
RMSEP.seasonal_factor

``` 
# 2 - Workings for Cosines harmonic model
```{r}
plot.ts(carbon.ts[1:8],main="First Seasons of carbon dioxide from  Jan 2000 - Dec 2001",xlab = "quarters", ylab="carbon dioxide (ppm)")
abline(v=2,lty=2)

Seasonal = cos((2*pi*(Time - 2))/4)
carbon.fit_cs_harmonic = lm(carbon.ts~Time+Seasonal)
#plot.ts(residuals(carbon.fit_cs_harmonic),main="Residual series of cosine harmonic")
#summary(carbon.fit_cs_harmonic)

carbon.fit_cs_harmonic2 = lm(carbon.ts~Time + I(Time^2) + Seasonal)
#plot.ts(residuals(carbon.fit_cs_harmonic2),main="Residual series of cosine harmonic with quadratic")

#summary(carbon.fit_cs_harmonic2)
#acf(residuals(carbon.fit_cs_harmonic2))

carbon.fit_cs_harmonic3 = lm(carbon.ts[-1]~Time[-1] + I(Time[-1]^2) + Seasonal[-1] + carbon.ts[-67])
acf(residuals(carbon.fit_cs_harmonic3))
summary(carbon.fit_cs_harmonic3)

normcheck(residuals(carbon.fit_cs_harmonic3),shapiro.wilk = T)
#y = carbon.fit_cs_harmonic3$coefficients[1] + carbon.fit_cs_harmonic3$coefficients[2] * 
#y = 811.3 + 1.024 * Time[-1] + 0.00190 * (Time[-1]^2) + 4.513 * cos(2*pi*(Time[-1]-2)/4) - 1.203
#y = 811.3 + 1.024 * Time + 0.00190 * (Time^2) + 4.513 * cos(2*pi*(Time-2)/4) - 1.203
#plot(Time[-1], y, type="l",main="Fitted Model")
#points(Time,carbon.ts, type = "p", col = "red")

#Forecast for Cosine
t68h.pred = carbon.fit_cs_harmonic3$coefficients[1] + carbon.fit_cs_harmonic3$coefficients[2] *68 + carbon.fit_cs_harmonic3$coefficients[3] * (68^2) - carbon.fit_cs_harmonic3$coefficients[4]*cos((2*pi*(68-2))/4) + carbon.fit_cs_harmonic3$coefficients[5] * carbon.ts[67]
t68h.pred

t69h.pred = carbon.fit_cs_harmonic3$coefficients[1] + carbon.fit_cs_harmonic3$coefficients[2] *69 + carbon.fit_cs_harmonic3$coefficients[3] * (69^2) - carbon.fit_cs_harmonic3$coefficients[4]*cos((2*pi*(69-2))/4) + carbon.fit_cs_harmonic3$coefficients[5] * t68h.pred
t69h.pred

t70h.pred = carbon.fit_cs_harmonic3$coefficients[1] + carbon.fit_cs_harmonic3$coefficients[2] *70 + carbon.fit_cs_harmonic3$coefficients[3] * (70^2) - carbon.fit_cs_harmonic3$coefficients[4]*cos((2*pi*(70-2))/4) + carbon.fit_cs_harmonic3$coefficients[5] * t69h.pred
t70h.pred

t71h.pred = carbon.fit_cs_harmonic3$coefficients[1] + carbon.fit_cs_harmonic3$coefficients[2] *71 + carbon.fit_cs_harmonic3$coefficients[3] * (71^2) - carbon.fit_cs_harmonic3$coefficients[4]*cos((2*pi*(71-2))/4) + carbon.fit_cs_harmonic3$coefficients[5] * t70h.pred
t71h.pred

print("The RMSEP")
RMSEP.cs_harmonic = sqrt((1/4)*sum((q_dioxide.df$dioxide[68]-t68h.pred)^2,(q_dioxide.df$dioxide[69]-t69h.pred)^2,(q_dioxide.df$dioxide[70]-t70h.pred)^2,(q_dioxide.df$dioxide[71]-t71h.pred)^2))
RMSEP.cs_harmonic

```

# 3- Full Harmonic Model
```{r}
c1 = cos(2*pi*Time*(1/4))
s1 = sin(2*pi*Time*(1/4))
c2 = cos(2*pi*Time*(2/4))

carbon.fit_fullH = lm(carbon.ts~Time+c1+s1+c2)
plot.ts(residuals(carbon.fit_fullH), main = "Residual Series of Full Harmonic")

carbon.fit_fullH2 = lm(carbon.ts~Time+I(Time^2)+c1+s1+c2)
plot.ts(residuals(carbon.fit_fullH2), main = "Residual Series of Full Harmonic with quadratic")

summary(carbon.fit_fullH2)
acf(residuals(carbon.fit_fullH))

carbon.fit_fullH3 = lm(carbon.ts[-1]~Time[-1]+I(Time[-1]^2)+c1[-1]+s1[-1]+c2[-1]+carbon.ts[-67])
summary(carbon.fit_fullH3)

#remove any non-significant variables (doesn't have to be a pair)
carbon.fit_fullH4 = lm(carbon.ts[-1]~Time[-1]+I(Time[-1]^2)+s1[-1]+c2[-1]+carbon.ts[-67])
summary(carbon.fit_fullH4)
acf(residuals(carbon.fit_fullH4))
plot.ts(residuals(carbon.fit_fullH4), main = "Residual Series")
normcheck(residuals(carbon.fit_fullH4), shapiro.wilk = T)
```
```{r}

t68.predFH = carbon.fit_fullH4$coefficients[1] + carbon.fit_fullH4$coefficients[2] *68 
+ carbon.fit_fullH4$coefficients[3] *(68^2) + carbon.fit_fullH4$coefficients[4]*sin(2*pi*68*(1/4)) + 
  carbon.fit_fullH4$coefficients[5]*cos(2*pi*68*(2/4)) + carbon.fit_fullH4$coefficients[6] * carbon.ts[67]
t68.predFH

t68.predFH2 = carbon.fit_fullH3$coefficients[1] + carbon.fit_fullH3$coefficients[2] *68 
+ carbon.fit_fullH3$coefficients[3] * (68^2) + carbon.fit_fullH3$coefficients[4]*cos(2*pi*68*(1/4)) + carbon.fit_fullH3$coefficients[5]*sin(2*pi*68*(1/4)) + 
  carbon.fit_fullH3$coefficients[6]*cos(2*pi*68*(2/4)) + carbon.fit_fullH3$coefficients[7]*carbon.ts[67]
#t68.predFH2 

#TO BE CONTINUED...
```


# 4 - Seasonally Adjusted Moving Averages Model 
```{r}
#decomp.series = decompose(carbon.ts)
#plot.ts(decomp.series, main = "ma_de-seasonalised_data")

decomp.ma.carbon = decompose(carbon.ts)
decomp.ma.carbon$dioxide
decomp.ma.carbon$figure
plot(decomp.ma.carbon)

decomp.ma.carbon$seasonal

ma.sa.carbon.ts = carbon.ts - decomp.ma.carbon$seasonal

plot(ma.sa.carbon.ts, main = "Moving average seasonally adjusted Carbon Dioxide" , xlab="Quarter", ylab="Carbon dioxide (ppm)")

carbon.fit_SA_MA = lm(ma.sa.carbon.ts~Time+ I(Time^2))
summary(carbon.fit_SA_MA)

plot.ts(residuals(carbon.fit_SA_MA), main = "Residual Series")
acf(residuals(carbon.fit_SA_MA))

carbon.fit_SA_MA2 = lm(ma.sa.carbon.ts[-1]~Time[-1]+ I(Time[-1]^2) + ma.sa.carbon.ts[-67])
summary(carbon.fit_SA_MA2)
acf(residuals(carbon.fit_SA_MA2))
normcheck(residuals(carbon.fit_SA_MA2), shapiro.wilk = T)


#SA MA Forecast

t68.ma.pred = carbon.fit_SA_MA2$coefficients[1] + carbon.fit_SA_MA2$coefficients[2]*68 + carbon.fit_SA_MA2$coefficients[3] *(68^2) + carbon.fit_SA_MA2$coefficients[4] * ma.sa.carbon.ts[67]
t68.ma.pred

t69.ma.pred = carbon.fit_SA_MA2$coefficients[1] + carbon.fit_SA_MA2$coefficients[2]*69 + carbon.fit_SA_MA2$coefficients[3] *(69^2) + carbon.fit_SA_MA2$coefficients[4] * t68.ma.pred
t69.ma.pred

t70.ma.pred = carbon.fit_SA_MA2$coefficients[1] + carbon.fit_SA_MA2$coefficients[2]*70 + carbon.fit_SA_MA2$coefficients[3] *(70^2) + carbon.fit_SA_MA2$coefficients[4] * t69.ma.pred
t70.ma.pred

t71.ma.pred = carbon.fit_SA_MA2$coefficients[1] + carbon.fit_SA_MA2$coefficients[2]*71 + carbon.fit_SA_MA2$coefficients[3] *(71^2) + carbon.fit_SA_MA2$coefficients[4] * t70.ma.pred
t71.ma.pred

print("The RMSEP")
RMSEP.ma.sa = sqrt((1/4)*sum((q_dioxide.df$dioxide[68]-t68.ma.pred)^2,(q_dioxide.df$dioxide[69]-t69.ma.pred)^2,(q_dioxide.df$dioxide[70]-t70.ma.pred)^2,(q_dioxide.df$dioxide[71]-t71.ma.pred)^2))
RMSEP.ma.sa

```


# 5 - Seasonally Adjusted STL Model
```{r}
decomp.stl.carbon = stl(carbon.ts, s.window = "periodic")
decomp.stl.carbon$time.series[1:4,1]
#plot(decomp.stl.carbon)

stl.sa.carbon.ts = carbon.ts-decomp.stl.carbon$time.series[,1]
plot(stl.sa.carbon.ts, main = "Seasonal Trend Lowess SA Carbon Dioxide", xlab="Quarter", ylab="Carbon Dioxide (ppm)")

carbon.fit_SA_STL = lm(stl.sa.carbon.ts[-1]~Time[-1]+I(Time[-1]^2) + stl.sa.carbon.ts[-67])
summary(carbon.fit_SA_STL)

plot.ts(residuals(carbon.fit_SA_STL), main = "Residual Series")
acf(residuals(carbon.fit_SA_STL))
normcheck(residuals(carbon.fit_SA_STL))


#STL Forecast 
t68.stl.pred = carbon.fit_SA_STL$coefficients[1] + carbon.fit_SA_STL$coefficients[2]*68 + carbon.fit_SA_STL$coefficients[3] *(68^2) + carbon.fit_SA_STL$coefficients[4] * stl.sa.carbon.ts[67]
t68.ma.pred

t69.stl.pred = carbon.fit_SA_STL$coefficients[1] + carbon.fit_SA_STL$coefficients[2]*69 + carbon.fit_SA_STL$coefficients[3] *(69^2) + carbon.fit_SA_STL$coefficients[4] * t68.stl.pred
t69.ma.pred

t70.stl.pred = carbon.fit_SA_STL$coefficients[1] + carbon.fit_SA_STL$coefficients[2]*70 + carbon.fit_SA_STL$coefficients[3] *(70^2) + carbon.fit_SA_STL$coefficients[4] * t69.stl.pred
t70.ma.pred

t71.stl.pred = carbon.fit_SA_STL$coefficients[1] + carbon.fit_SA_STL$coefficients[2]*71 + carbon.fit_SA_STL$coefficients[3] *(71^2) + carbon.fit_SA_STL$coefficients[4] * t70.stl.pred
t71.ma.pred

print("The RMSEP")
RMSEP.stl.sa = sqrt((1/4)*sum((q_dioxide.df$dioxide[68]-t68.stl.pred)^2,(q_dioxide.df$dioxide[69]-t69.stl.pred)^2,(q_dioxide.df$dioxide[70]-t70.stl.pred)^2,(q_dioxide.df$dioxide[71]-t71.stl.pred)^2))
RMSEP.stl.sa

```



#Question 3 - 

Re-running Seasonally adjusted Moving averages model - with all data
```{r}
QuarterlyDioxide.ts = ts(qdioxide$dioxide, frequency = 4, start = 2000)
Time.sa = 1:71
Quarter.sa = factor(c(rep(1:4,17),(1:3)))
carbon.sa.ts = ts(q_dioxide.df$dioxide[1:71],start=2000, frequency=4)


decomp.ma.carbon3 = decompose(carbon.sa.ts)
decomp.ma.carbon3$dioxide
decomp.ma.carbon3$figure
plot(decomp.ma.carbon3)

decomp.ma.carbon3$seasonal

ma.sa.carbon3.ts = carbon.sa.ts - decomp.ma.carbon3$seasonal

plot(ma.sa.carbon3.ts, main = "Seasonally Adjusted Moving Averages Carbon Dioxide from Quarter 1 (2000) - Quarter 3 (2017) " , xlab="Quarter", ylab="Carbon Dioxide (ppm)")

carbon.fit3 = lm(ma.sa.carbon3.ts~Time.sa+ I(Time.sa^2))
summary(carbon.fit3)

plot.ts(residuals(carbon.fit3), main = "Residual Series")
acf(residuals(carbon.fit3))

carbon.fit4 = lm(ma.sa.carbon3.ts[-1]~Time.sa[-1]+ I(Time.sa[-1]^2) + ma.sa.carbon3.ts[-71])
summary(carbon.fit4)
acf(residuals(carbon.fit4))
normcheck(residuals(carbon.fit4), shapiro.wilk = T)


#SA MA Forecast

t72.ma.pred = carbon.fit4$coefficients[1] + carbon.fit4$coefficients[2]*72 + carbon.fit4$coefficients[3] *(72^2) + carbon.fit4$coefficients[4] * ma.sa.carbon3.ts[71]
t72.ma.pred

t73.ma.pred = carbon.fit4$coefficients[1] + carbon.fit4$coefficients[2]*73 + carbon.fit4$coefficients[3] *(73^2) + carbon.fit4$coefficients[4] * t72.ma.pred
t73.ma.pred

t74.ma.pred = carbon.fit4$coefficients[1] + carbon.fit4$coefficients[2]*74 + carbon.fit4$coefficients[3] *(74^2) + carbon.fit4$coefficients[4] * t73.ma.pred
t74.ma.pred

t75.ma.pred = carbon.fit4$coefficients[1] + carbon.fit4$coefficients[2]*75 + carbon.fit4$coefficients[3] *(75^2) + carbon.fit4$coefficients[4] * t74.ma.pred
t75.ma.pred

#TO DO PREDICTION INTERVALS 

pred72.high = t72.ma.pred + 1.96 * 0.3452
pred72.low = t72.ma.pred - 1.96 * 0.3452

cat("Lower t72: ",pred72.low, "\n")
cat("Upper t72: ",pred72.high, "\n")


pred73.high = t73.ma.pred + 1.96 * 0.3452
pred73.low = t73.ma.pred - 1.96 * 0.3452

cat("Lower t73: ",pred73.low, "\n")
cat("Upper t73: ",pred73.high, "\n")


pred74.high = t74.ma.pred + 1.96 * 0.3452
pred74.low = t74.ma.pred - 1.96 * 0.3452

cat("Lower t74: ",pred74.low, "\n")
cat("Upper t74: ",pred74.high, "\n")


pred75.high = t75.ma.pred + 1.96 * 0.3452
pred75.low = t75.ma.pred - 1.96 * 0.3452

cat("Lower t75: ",pred75.low, "\n")
cat("Upper t75: ",pred75.high, "\n")
```
Technical notes for question 3:





#Question 4 - Holt Winters Model of Q3
```{r}
#Additive Model HS

carbon.fit.hw = HoltWinters(carbon.ts)
plot(carbon.fit.hw)

carbon.fit.hw.pred = predict(carbon.fit.hw, n.ahead = 4)
#carbon.fit.hw.pred

plot(carbon.fit.hw,carbon.fit.hw.pred)

actual = q_dioxide.df$dioxide[68:71]
actual

RMSEP.hw.add.pred = sqrt(1/4 * sum((actual-carbon.fit.hw.pred)^2))
RMSEP.hw.add.pred
```
Comments: justify answer for this being better = RMSEP is smaller

